---
title: "2022 Lagoons Preliminary Data Quality Assessment"
author: "SDMiller"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(readxl)
library(tidyverse)
library(sqldf)
library(lubridate)

connection = dbConnect(odbc(),Driver = "Sql Server",Server = "inpyugamsvm01\\nuna_dev", Database = "ARCN_Lagoons")

dir = r'(C:\Work\VitalSigns\ARCN Lagoons\Data\2022 Lagoons Sampling\Data processing/)' # copied local for speed
workbook = paste(dir,'CAKR 2022 fish and water quality data.xlsx',sep="")
Year=2022

```

# 2022 Lagoons data ETL

# Functions
```{r}
# Example database connection
# connection = dbConnect(odbc(),Driver = "Sql Server",Server = "inpyugamsvm01\\nuna_dev", Database = "ARCN_Lagoons")

# Function to query the lagoons database for the existence of Lagoon using the odbc database connection DatabaseConnection
# Returns boolean.
LagoonExists = function(DatabaseConnection, Lagoon) {
  tryCatch({
    # Build an SQL query
    Sql = paste("SELECT Lagoon FROM Lagoons WHERE Lagoon = '", Lagoon, "'", sep = "")
    
    # Execute the query
    result = dbGetQuery(DatabaseConnection, Sql)
    
    # Return true or false depending on whether we get a row or not
    if (nrow(result) == 0) {
      return(paste("INSERT INTO Lagoons(Lagoon)VALUES('",Lagoon,"') -- ",Lagoon," Does not exist\n",sep=""))
    } else {
      return(paste(Sql," -- Exists\n",sep=""))
    }
  }, error = function(e) {
    # Bummer
    cat("Error: ", e$message,"\n",Sql,"\n")
    return(NULL) # or handle the error as appropriate
  })
}
# Test cat(LagoonExists(connection,'Anigaaqi'))

# Function to query the lagoons database for the existence of Lagoon and Site using the odbc database connection DatabaseConnection
# Returns boolean.
SiteExists = function(DatabaseConnection, Lagoon,Site,Lat='NULL',Lon='NULL') {
  tryCatch({
    # Build an SQL query
    Sql = paste("SELECT Lagoon,Site FROM Sites WHERE Lagoon = '", Lagoon, "' And Site = '",Site,"'", sep = "")

    # Execute the query
    result = dbGetQuery(DatabaseConnection, Sql)
    
    # Return true or false depending on whether we get a row or not
    if (nrow(result) == 0) {
      
      if(is.na(Lat) | Lat == 0){
        return(paste("INSERT INTO Sites(Lagoon,Site,Latitude,Longitude)VALUES('",Lagoon,"','",Site,"',NULL,NULL)",sep=""))
      } else {
        return(paste("INSERT INTO Sites(Lagoon,Site,Latitude,Longitude)VALUES('",Lagoon,"','",Site,"',",Lat,",",Lon,")",sep=""))
      }

    } else {
      return(paste(Sql," -- Exists\n",sep=""))
    }
    
  }, error = function(e) {
    # Bummer
    cat("Error: ", e$message,"\n",Sql,"\n")
    return(NULL) # or handle the error as appropriate
  })
}
# cat(SiteExists(connection,'Anigaaq','ANI_ME',65,-157),sep="")
# cat(SiteExists(connection,'Anigaaq','ANI_ME',0,0),sep="")
# cat(SiteExists(connection,'Anigaaq','ANI_ME',NA,NA),sep="")
# cat(SiteExists(connection,'Anigaaq','ANI_NS'),sep="")


# Function to query the lagoons database for the existence of a lagoons sampling event using the odbc database connection DatabaseConnection
# Returns boolean.
SamplingEventExists = function(DatabaseConnection, Lagoon,Date,Site) {
  tryCatch({
    # Build an SQL query
    Sql = paste("SELECT Lagoon,[Date],Site FROM SamplingEvents WHERE Lagoon = '", Lagoon, "' And [Date] = '",Date,"' And Site = '",Site,"'", sep = "")

    # Execute the query
    result = dbGetQuery(DatabaseConnection, Sql)
    
    
    # Return true or false depending on whether we get a row or not
    if (nrow(result) == 0) {
        return(paste("INSERT INTO SamplingEvents(Lagoon,Site,Date)VALUES('",Lagoon,"','",Site,"','",Date,"')",sep=""))
    } else {
      return(paste(Sql," -- Exists\n",sep=""))
    }
    
    
  }, error = function(e) {
    # Bummer
    cat("Error: ", e$message,"\n",Sql,"\n")
    return(NULL) # or handle the error as appropriate
  })
}
# cat(SamplingEventExists(connection,'Anigaaq','2024-06-23','ANI_ME'))
# cat(SamplingEventExists(connection,'Aukulak','2024-06-19','AUK_OUT'))

# Summarizes the parameters of a data frame
GetDataFrameSummary = function(DataFrame){
    # Calculate min, max, and mean for numeric columns
  numeric_summary <- DataFrame %>%  summarise(across(where(is.numeric), list(Min = ~min(., na.rm = TRUE),Max = ~max(., na.rm = TRUE),Mean = ~mean(., na.rm = TRUE))))
  
  # Reshape the summary statistics to have parameters as rows and stats as columns
  numeric_summary_long <- pivot_longer(numeric_summary, everything(), names_to = c("Parameter", ".value"), names_sep = "_")
}

```


# Data file to ARCN_Lagoons database processing steps



# Fish

## Load fish data
```{r}

# Get the sheets: excel_sheets(xlfile)
cat(xlfile,"\n")
Worksheet = paste("Fish data",sep="")

# Fish data
fd = read_excel(xlfile,Worksheet)
fd$Date = as.Date(fd$Date) # dates to real dates

glimpse(fd)
```

## Lagoons exist check
```{r}

# Check lagoons exist
Lagoons  = fd %>% distinct(Lagoon) %>% arrange(Lagoon)
for (i in 1:nrow(Lagoons)){
  Lagoon = unlist(Lagoons[i,'Lagoon'])
  cat(Lagoon," exists: ",LagoonExists(connection,Lagoon),"\n")
}

```

## Sites exist check
```{r}

# Check sites exist
Sites  = fd %>% distinct(Lagoon,Site) %>% arrange(Lagoon,Site)
for (i in 1:nrow(Sites)){
  Lagoon = unlist(Sites[i,'Lagoon'])
  Site = unlist(Sites[i,'Site'])
  cat(i," ",Lagoon,"-",Site," exists: ",SiteExists(connection,Lagoon,Site),"\n")
}

```

## Sampling events exist
```{r}
# Check sampling events exist
Events  = fd %>% distinct(Lagoon,Site,Date) %>% arrange(Lagoon,Site,Date)
Events$DateText = as.character(Events$Date)
cat("begin transaction -- commit rollback\n")
for (i in 1:nrow(Events)){
  Lagoon = unlist(Events[i,'Lagoon'])
  Site = unlist(Events[i,'Site'])
  Date = unlist(Events[i,'DateText'])
  Exists = SamplingEventExists(connection,Lagoon,Date,Site)
  
  if(Exists==TRUE){
    cat("-- ",i," ",Lagoon,"-",Site,", ",Date," exists: ",Exists,"\n",sep="")
  }else{
    cat("INSERT INTO SamplingEvents(Lagoon,Site,StartDateTime)VALUES('",Lagoon,"','",Site,"','",Date,"'); -- ",i,"\n",sep="")
  }
}

```

## Fix problems

```{r}
# unmatched species
# fd %>% filter(is.na(CommonName)==TRUE) %>% distinct(CommonName)
# fd$SetTime_Min = as.integer(fd$`Set Time`)
# 
# glimpse(fd)



```

## Step 4. Join the fish scientific names
```{r}
Sql = "select distinct CommonName,Family,Genus,Species,CommonName as FBname from fish where CommonName is not null order by CommonName,Family,Genus,Species"
Sql = "SELECT * from fishbase"
fb = dbGetQuery(connection,Sql)
fb$CommonNameUpper = toupper(fb$FBname)
fd$CommonNameUpper = toupper(fd$CommonName)
x = as.data.frame(left_join(fd,fb,by='CommonNameUpper',multiple='first'))

x %>% filter(Family=='' | is.na(Family)) %>% distinct(CommonNameUpper)

# glimpse(x)

### juvenile salmon
x$Family = ifelse(toupper(x$CommonName)=='JUVENILE SALMON','Salmonidae',x$Family)
#x$Genus = ifelse(toupper(x$CommonName)=='JUVENILE SALMON','Oncorhynchus',x$Genus)
#x$Species = ifelse(toupper(x$CommonName)=='JUVENILE SALMON','keta',x$Species)

### pighead (st)pricklback
CommonName = 'PIGHEAD STICKLEBACK'
x$Family = ifelse(toupper(x$CommonName)=='PIGHEAD STICKLEBACK','Lumpenidae',x$Family)
x$Genus = ifelse(toupper(x$CommonName)=='PIGHEAD STICKLEBACK','Acantholumpenus',x$Genus)
x$Species = ifelse(toupper(x$CommonName)=='PIGHEAD STICKLEBACK','mackayi',x$Species)
x$CommonName = ifelse(toupper(x$CommonName)=='PIGHEAD STICKLEBACK','Pighead prickleback',x$CommonName)
```


## Step 5. Write the fish deliverable

```{r}
# glimpse(x)
DeliverableFile = paste(dir,"L11 2022 Lagoons Fish Data.csv",sep="")
write.csv(x,DeliverableFile,row.names = FALSE,na="")

```


















# 2022 Deliverable L10-Discrete water quality data processing

Things to do in Excel
* Convert weird Excel integer dates to date text values, =TEXT(A2,"YYYY-MM-DD")
* Fix non-standard Site values. Example =UPPER(LEFT(A2,3))&"_ME" yields AUK_ME for Aukulak Marine Edge
* Move Site values to SiteRecorded
* Blankify NAs
* Anything else that's easier to do in Excel  than R


## Step 1. Load the worksheet

```{r}
# Get the worksheet names
#excel_sheets(workbook)

# Load the fish data
worksheet = "Water quality"
wd = readxl::read_excel(workbook,worksheet)
# Use these tools to check that data types match expected attributes, statistics such as min/max, etc. are as expected
wd$Date = as.Date(wd$Date)

# glimpse(wd)

# summary(wd)

```

## Step 2. Data quality checks

```{r}

# Generate dplyr distinct values statements
for(ColumnName in colnames(wd)){
  cat("wd %>% distinct(`",ColumnName,"`) %>% arrange(`",ColumnName,"`)\n",sep="")
}

```

```{r}

wd %>% distinct(`Date`) %>% arrange(`Date`)
wd %>% distinct(`Lagoon`) %>% arrange(`Lagoon`)
wd %>% distinct(`Site`) %>% arrange(`Site`)
wd %>% distinct(`SiteType`) %>% arrange(`SiteType`)
wd %>% distinct(`Location`) %>% arrange(`Location`)
wd %>% distinct(`TEMP (Celsius)`) %>% arrange(`TEMP (Celsius)`)
wd %>% distinct(`SPC`) %>% arrange(`SPC`)
wd %>% distinct(`Salinity`) %>% arrange(`Salinity`)
wd %>% distinct(`ODO%`) %>% arrange(`ODO%`)
wd %>% distinct(`pH`) %>% arrange(`pH`)
wd %>% distinct(`phFlag`) %>% arrange(`phFlag`)
wd %>% distinct(`Chlorophyl RFU`) %>% arrange(`Chlorophyl RFU`)
wd %>% distinct(`BGA RFU`) %>% arrange(`BGA RFU`)
wd %>% distinct(`Turbidity (FNU)`) %>% arrange(`Turbidity (FNU)`)
wd %>% distinct(`ExcelDate`) %>% arrange(`ExcelDate`)

```
### 2022 Lagoons existence check

```{r}
# fix lagoon names
# wd$Lagoon = ifelse(wd$Lagoon=='Aniqaaq','Anigaaq',wd$Lagoon)
# wd$Lagoon = ifelse(wd$Lagoon=='Aukulaaq','Aukulak',wd$Lagoon)


# make sure lagoons exist
Lagoons = as.data.frame(wd %>% distinct(Lagoon))

# Rabbit Creek is not a lagoon
for(i in 1:nrow(Lagoons)){
  Lagoon = Lagoons[i,1]
  cat(LagoonExists(connection,Lagoon))
  #cat("INSERT INTO Lagoons(Lagoon)VALUES('",Lagoon,"') -- ",Lagoon," ",Exists,"\n",sep="")
}

```

### 2022 Sites existence check

```{r}
wd$Lat = NA
wd$Long = NA
# Check sites exist, if not get insert queries
Sites = wd %>% distinct(Lagoon,Site,Lat,Long)
Sites
cat("BEGIN TRANSACTION -- COMMIT ROLLBACK\n")
for(i in 1:nrow(Sites)){
  Lagoon = Sites[i,'Lagoon']
  Site = Sites[i,'Site']
  Lat = Sites[i,'Lat']
  Lon = Sites[i,'Long']
  cat(SiteExists(connection,Lagoon,Site,Lat,Lon)," -- ",i,"\n",sep="")
}

```



### 2022 Sampling Events existence check

```{r}
Events = wd %>% distinct(Lagoon,Date,Site) %>% arrange(Site)
Events$RealDate = as.character(Events$Date)

cat("BEGIN TRANSACTION -- COMMIT ROLLBACK\n")
cat("-- ",nrow(Events)," rows\n")
for(i in 1:nrow(Events)){
  Lagoon = Events[i,'Lagoon']
  Site = Events[i,'Site']
  Date = Events[i,'RealDate']
  cat(SamplingEventExists(connection,Lagoon,Date,Site),"-- ",i,"\n")
}

```

## Step 5. Write the water quality deliverable

```{r}

write.csv(wd,paste(dir,"L10 2022 Lagoons Discrete Water Quality Data.csv",sep=""),row.names = FALSE,na="")

```




