---
title: "2021/2024 Lagoons Data ETL"
author: "SDMiller"
date: "`r Sys.Date()`"
output: html_document
---

# Intro

This document describes the steps I used to implement quality control on the data deliverables for 2021/2024 for ingestion into the ARCN_Lagoons database.

These data were received from WCS, Kevin Fraley.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
library(sqldf)
library(tidyverse)
library(readxl)
library(odbc)
# library(leaflet)

connection = dbConnect(odbc(),Driver = "Sql Server",Server = "inpyugamsvm01\\nuna_dev", Database = "ARCN_Lagoons")
dir = r'(O:\Monitoring\Vital Signs\Lagoon Communities and Ecosystems\Data\2024 Lagoons Sampling\Deliverables processing/)'
dir = r'(C:\Work\VitalSigns\ARCN Lagoons\Data\2024 Lagoons Sampling\Deliverables processing/)' # copied local for speed
filename = r'(2021-2024 WCS CAKR Coastal Lagoons Data Received 2024-07-31 From KFraley.xlsx)'
workbook = paste(dir,filename,sep="")

```

# Functions
```{r}
# Example database connection
# connection = dbConnect(odbc(),Driver = "Sql Server",Server = "inpyugamsvm01\\nuna_dev", Database = "ARCN_Lagoons")

# Function to query the lagoons database for the existence of Lagoon using the odbc database connection DatabaseConnection
# Returns boolean.
LagoonExists = function(DatabaseConnection, Lagoon) {
  tryCatch({
    # Build an SQL query
    Sql = paste("SELECT Lagoon FROM Lagoons WHERE Lagoon = '", Lagoon, "'", sep = "")
    
    # Execute the query
    result = dbGetQuery(DatabaseConnection, Sql)
    
    # Return true or false depending on whether we get a row or not
    if (nrow(result) == 0) {
      return(paste("INSERT INTO Lagoons(Lagoon)VALUES('",Lagoon,"') -- ",Lagoon," Does not exist\n",sep=""))
    } else {
      return(paste(Sql," -- Exists\n",sep=""))
    }
  }, error = function(e) {
    # Bummer
    cat("Error: ", e$message,"\n",Sql,"\n")
    return(NULL) # or handle the error as appropriate
  })
}
# Test cat(LagoonExists(connection,'Anigaaqi'))

# Function to query the lagoons database for the existence of Lagoon and Site using the odbc database connection DatabaseConnection
# Returns boolean.
SiteExists = function(DatabaseConnection, Lagoon,Site,Lat='NULL',Lon='NULL') {
  tryCatch({
    # Build an SQL query
    Sql = paste("SELECT Lagoon,Site FROM Sites WHERE Lagoon = '", Lagoon, "' And Site = '",Site,"'", sep = "")

    # Execute the query
    result = dbGetQuery(DatabaseConnection, Sql)
    
    # Return true or false depending on whether we get a row or not
    if (nrow(result) == 0) {
      
      if(is.na(Lat) | Lat == 0){
        return(paste("INSERT INTO Sites(Lagoon,Site,Latitude,Longitude)VALUES('",Lagoon,"','",Site,"',NULL,NULL)",sep=""))
      } else {
        return(paste("INSERT INTO Sites(Lagoon,Site,Latitude,Longitude)VALUES('",Lagoon,"','",Site,"',",Lat,",",Lon,")",sep=""))
      }

    } else {
      return(paste(Sql," -- Exists\n",sep=""))
    }
    
  }, error = function(e) {
    # Bummer
    cat("Error: ", e$message,"\n",Sql,"\n")
    return(NULL) # or handle the error as appropriate
  })
}
# cat(SiteExists(connection,'Anigaaq','ANI_ME',65,-157),sep="")
# cat(SiteExists(connection,'Anigaaq','ANI_ME',0,0),sep="")
# cat(SiteExists(connection,'Anigaaq','ANI_ME',NA,NA),sep="")
# cat(SiteExists(connection,'Anigaaq','ANI_NS'),sep="")


# Function to query the lagoons database for the existence of a lagoons sampling event using the odbc database connection DatabaseConnection
# Returns boolean.
SamplingEventExists = function(DatabaseConnection, Lagoon,Date,Site) {
  tryCatch({
    # Build an SQL query
    Sql = paste("SELECT Lagoon,[Date],Site FROM SamplingEvents WHERE Lagoon = '", Lagoon, "' And [Date] = '",Date,"' And Site = '",Site,"'", sep = "")

    # Execute the query
    result = dbGetQuery(DatabaseConnection, Sql)
    
    
    # Return true or false depending on whether we get a row or not
    if (nrow(result) == 0) {
        return(paste("INSERT INTO SamplingEvents(Lagoon,Site,Date)VALUES('",Lagoon,"','",Site,"','",Date,"')",sep=""))
    } else {
      return(paste(Sql," -- Exists\n",sep=""))
    }
    
    
  }, error = function(e) {
    # Bummer
    cat("Error: ", e$message,"\n",Sql,"\n")
    return(NULL) # or handle the error as appropriate
  })
}
# cat(SamplingEventExists(connection,'Anigaaq','2024-06-23','ANI_ME'))
# cat(SamplingEventExists(connection,'Aukulak','2024-06-19','AUK_OUT'))

# Summarizes the parameters of a data frame
GetDataFrameSummary = function(DataFrame){
    # Calculate min, max, and mean for numeric columns
  numeric_summary <- DataFrame %>%  summarise(across(where(is.numeric), list(Min = ~min(., na.rm = TRUE),Max = ~max(., na.rm = TRUE),Mean = ~mean(., na.rm = TRUE))))
  
  # Reshape the summary statistics to have parameters as rows and stats as columns
  numeric_summary_long <- pivot_longer(numeric_summary, everything(), names_to = c("Parameter", ".value"), names_sep = "_")
}

```

# 2021 Deliverable L11-Fish data processing

This section contains processing steps to get the fish data (Deliverable L11) processed for ingestion into the master lagoons monitoring database.

Preprocessing in Excel
* Site names standardization
* NA removal
* Fix Excel dates

## Step 1. Load the worksheet

```{r}
# Get the worksheet names
#excel_sheets(workbook)

# Load the fish data
worksheet = "Fish Data 2021"
fd = readxl::read_excel(workbook,worksheet)

```

## Step 2. Data quality checks

```{r}

# Use these tools to check that data types match expected attributes, statistics such as min/max, etc. are as expected
fd$Date = as.Date(fd$Date)

glimpse(fd)
Summary = GetDataFrameSummary(fd)
knitr::kable(Summary)


# Generate dplyr distinct values statements
for(ColumnName in colnames(fd)){
  cat("fd %>% distinct(",ColumnName,") %>% arrange(",ColumnName,")\n",sep="")
}

fd$SetTime_Min = fd$`Set Time (hours)`*60

# Look at distinct values
fd %>% distinct(Year) %>% arrange(Year)
fd %>% distinct(Date) %>% arrange(Date)
fd %>% distinct(Lagoon) %>% arrange(Lagoon)
fd %>% distinct(Site) %>% arrange(Site)
fd %>% distinct(SiteType) %>% arrange(SiteType)
fd %>% distinct(SiteRecorded) %>% arrange(SiteRecorded)
fd %>% distinct(GearType) %>% arrange(GearType)
fd %>% distinct(`Set Time (hours)`) %>% arrange(`Set Time (hours)`)
fd %>% distinct(`Check Number`) %>% arrange(`Check Number`)
fd %>% distinct(`Check Time`) %>% arrange(`Check Time`)
fd %>% distinct(CommonName) %>% arrange(CommonName)
fd %>% distinct(Count) %>% arrange(Count)
fd %>% distinct(ForkLength) %>% arrange(ForkLength)
fd %>% distinct(ForkLengthText) %>% arrange(ForkLengthText)
fd %>% distinct(Notes) %>% arrange(Notes)
fd %>% distinct(Lat) %>% arrange(Lat)
fd %>% distinct(Long) %>% arrange(Long)
fd %>% distinct(ExcelDate) %>% arrange(ExcelDate)



```

## Step 3. Fix defects, 2021

### Fix lagoons

```{r}
# fix lagoon names
# fd$Lagoon = ifelse(fd$Lagoon=='Aniqaaq','Anigaaq',fd$Lagoon)
# fd$Lagoon = ifelse(fd$Lagoon=='Aukulaaq','Aukulak',fd$Lagoon)

# offload non-existent lagoons
# write to a file and filter out of the dataset
# write.csv(fd %>% filter(Lagoon == 'Rabbit Creek'),paste(dir,"L11 2021 Rabbit Creek Fish Data.csv",sep=""))
# write.csv(fd %>% filter(Lagoon == 'Situkuyok'),paste(dir,"L11 2021 Situkuyok Fish Data.csv",sep=""))
# 
# fd = fd %>% filter(Lagoon != 'Rabbit Creek' & Lagoon != 'Situkuyok')
# i don't know where rabbit creek or Situkuyok is, no spatial data


```

### 2021 Lagoons existence check

```{r}

# make sure lagoons exist
Lagoons = as.data.frame(fd %>% distinct(Lagoon))

# Rabbit Creek is not a lagoon
for(i in 1:nrow(Lagoons)){
  Lagoon = Lagoons[i,1]
  cat(LagoonExists(connection,Lagoon))
  #cat("INSERT INTO Lagoons(Lagoon)VALUES('",Lagoon,"') -- ",Lagoon," ",Exists,"\n",sep="")
}

# ValidLagoonNames = dbGetQuery(connection,"SELECT DISTINCT Lagoon FROM Lagoons ORDER BY Lagoon")



```


### Fix Sites

```{r}


# Fix Site names
# in order to fix Site we need to move it's current data into SiteType
#fd$SiteType = fd$Site

# glimpse(fd)

# Distinct sites
# fd %>% distinct(Site,SiteType) %>% arrange(Site)

# Create site names
#fd$Site = paste(toupper(substr(fd$Lagoon,1,3)),"_",toupper(substr(fd$SiteType,1,3)), sep="")

# go back and fix items
#fd$Site = ifelse(fd$SiteType=='Inflow',paste(toupper(substr(fd$Lagoon,1,3)),"_IN", sep=""),fd$Site)
#fd$Site = ifelse(fd$SiteType=='Marine Edge',paste(toupper(substr(fd$Lagoon,1,3)),"_ME", sep=""),fd$Site)
#fd$Site = ifelse(fd$SiteType=='Marine',paste(toupper(substr(fd$Lagoon,1,3)),"_ME", sep=""),fd$Site)
#fd$Site = ifelse(fd$SiteType=='Marine edge',paste(toupper(substr(fd$Lagoon,1,3)),"_ME", sep=""),fd$Site)

```

### 2021 Sites existence check

```{r}

# Check sites exist, if not get insert queries
Sites = fd %>% distinct(Lagoon,Site,Lat,Long)
Sites
cat("BEGIN TRANSACTION -- COMMIT ROLLBACK\n")
for(i in 1:nrow(Sites)){
  Lagoon = Sites[i,'Lagoon']
  Site = Sites[i,'Site']
  Lat = Sites[i,'Lat']
  Lon = Sites[i,'Long']
  cat(SiteExists(connection,Lagoon,Site,Lat,Lon)," -- ",i,"\n",sep="")
}

```

### Fix Sampling Events

```{r}
# Fix Sites
Events = fd %>% distinct(Lagoon,Date,Site) %>% arrange(Site)
# Events$RealDate = as.character(Events$Date)

```

### 2021 Sampling Events existence check

```{r}
Events = fd %>% distinct(Lagoon,Date,Site) %>% arrange(Site)
Events$DateChar = as.character(Events$Date)
cat("BEGIN TRANSACTION -- COMMIT ROLLBACK\n")
cat("-- ",nrow(Events)," rows\n")

for(i in 1:nrow(Events)){
  Lagoon = Events[i,'Lagoon']
  Site = Events[i,'Site']
  Date = as.character(Events[i, 'DateChar'])
  cat(SamplingEventExists(connection,Lagoon,Date,Site),"\n",sep="")
}

```



### Fix Gear type
```{r}
# Move gear type to CollectionMethod to preserve gear type
fd$CollectionMethod = fd$`Gear Type`
fd$CollectionMethod = ifelse(fd$`Gear Type`=='offshore gillnet','Gillnet',fd$CollectionMethod)
fd$CollectionMethod = ifelse(fd$`Gear Type`=='onshore gillnet','Gillnet',fd$CollectionMethod)
fd$CollectionMethod = ifelse(fd$`Gear Type`=='beach Seine','Beach Seine',fd$CollectionMethod)
fd$CollectionMethod = ifelse(fd$`Gear Type`=='Onshore gillnet','Gillnet',fd$CollectionMethod)
fd$CollectionMethod = ifelse(fd$`Gear Type`=='Fike','Fyke',fd$CollectionMethod)

fd %>% distinct(`Gear Type`,CollectionMethod)

```

### Fix Set time
```{r}
fd %>% distinct(`Set Time (hours)`)
# These should be in minutes
fd$SetTime_Min = as.integer(fd$`Set Time (hours)`*60)

fd %>% distinct(`Set Time (hours)`,SetTime_Min)


```


### Fix Check Time
```{r}
fd %>% distinct(`Check Time`)
fd$CheckTime= ifelse(!is.na(fd$`Check Time`), paste(fd$`Check Time`,":00",sep=""),NA)

fd %>% distinct(CheckTime)

```
### Fix Species
```{r}
fd %>% distinct(Species)
fd$CommonName = fd$Species

fd$CommonName = ifelse(fd$Species=='No fish',NA,fd$CommonName)
fd$CommonName = ifelse(fd$Species=='No Fish',NA,fd$CommonName)
fd %>% distinct(CommonName)
```

### Fix Fork Length
```{r}

# Fork length is character because of censored data. Some values have a '+', for example
fd$ForkLength = ifelse(is.na(as.numeric(fd$`Fork Length (mm)`)), NA, as.numeric(fd$`Fork Length (mm)`))
# Move text fork lengths into a new column
fd$ForkLengthText = ifelse(is.na(as.numeric(fd$`Fork Length (mm)`)), fd$`Fork Length (mm)`, NA)
# Flag the text fork length data
fd$ForkLengthFlag = ifelse(!is.na(fd$ForkLengthText),6,NA)

# Get the numeric value of fork length text and load it into ForkLength
fd$ForkLength = ifelse(
  is.na(as.numeric(fd$`Fork Length (mm)`)),
  gsub("[^0-9]", "", fd$ForkLengthText)
  ,fd$ForkLength
)
#fd %>% filter(ForkLengthFlag == 6) %>% mutate(ForkLength = gsub("[^0-9]", "", ForkLengthText)) %>% select(ForkLength,ForkLengthText,ForkLengthFlag) 

fd$ForkLength = ifelse(fd$ForkLength == 150300,225,fd$ForkLength)

# check work
#fd %>% select(`Fork Length (mm)`,ForkLength,ForkLengthText,ForkLengthFlag) %>% arrange(ForkLength)
fd %>% distinct(ForkLength,`Fork Length (mm)`,ForkLengthText,ForkLengthFlag) %>% arrange(ForkLengthText)

```

### Fix Counts

```{r}
# Move the text counts to a new column
fd$Count_Text  = NULL

# Fix by moving the text values into a new column
fd$CountText = ifelse(is.na(as.numeric(fd$Count)), fd$Count, NA)

# Flag the text count data
fd$CountFlag = ifelse(!is.na(fd$CountText),7,NA)

# Get the numeric value of count and load it into Count
fd$Count = ifelse(
  is.na(as.numeric(fd$Count)),
  gsub("[^0-9]", "", fd$CountText)
  ,fd$Count
)

# check work
fd %>% distinct(CountText,Count,CountFlag) %>% arrange(CountText,Count)
fd %>% select(Lagoon,Site,Date,CommonName,Count,CountText,CountFlag) %>% arrange(Count,CountText,CountFlag,Lagoon,Site,Date,CommonName)

```
### Fix Lat Lon

```{r}
fd %>% distinct(Lat) %>% arrange(Lat)
fd %>% distinct(Long) %>% arrange(Long)
```



## Step 4. Join the fish scientific names
```{r}
fb = dbGetQuery(connection,"SELECT * from fishbase")
fb$CommonNameUpper = toupper(fb$FBname)
fd$CommonNameUpper = toupper(fd$CommonName)
x = as.data.frame(left_join(fd,fb,by='CommonNameUpper',multiple='first'))
# glimpse(x)

### Fix Species
# x %>% distinct(Genus,Species.y) %>% arrange(Genus,Species.y)

```


## Step 5. Write the fish deliverable

```{r}
glimpse(x)

deliverable = x %>% select(Date
,Lagoon
,Site
,Family
,Genus
,`Species`
,CommonName
,GearType
,`Check Number` 
,`Check Time`
,SetTime_Min
,Count
,ForkLength
,Lat
,Long
,Notes,SiteType)

# deliverable = sqldf("SELECT Date
# ,Lagoon
# ,Site
# ,Family
# ,Genus
# ,[Species.y] as Species
# ,CommonName
# ,CollectionMethod
# ,[Check Number] as CheckNumber
# ,CheckTime
# ,SetTime_Min
# ,Count,CountText,CountFlag
# ,ForkLength as ForkLength_mm,ForkLengthText,ForkLengthFlag as ForkLengthFlag
# ,Lat as Latitude
# ,Long as Longitude
# ,[Species.x] as CommonNameRecorded
# ,Notes as Comments,SiteType as Site_Recorded
# FROM x 
# ORDER BY Date
# ,Lagoon
# ,Site,CollectionMethod,CheckNumber")

write.csv(deliverable,paste(dir,"L11 2021 Lagoons Fish Data.csv",sep=""),row.names = FALSE,na="")

```




# 2024 Deliverable L11-Fish data processing

This section contains processing steps to get the fish data (Deliverable L11) processed for ingestion into the master lagoons monitoring database.

## Step 1. Load the worksheet

```{r}
# Get the worksheet names
#excel_sheets(workbook)

# Load the fish data
worksheet = "Fish Data 2024"
fd = readxl::read_excel(workbook,worksheet)

```

## Step 2. Data quality checks

```{r}

# Use these tools to check that data types match expected attributes, statistics such as min/max, etc. are as expected
glimpse(fd)

# fd %>% distinct(Year)
# fd %>% distinct(Date)
# fd %>% distinct(Lagoon)
# fd %>% distinct(Site)
# fd %>% distinct(`Gear Type`)
# fd %>% distinct(`Set Time (hours)`)
# fd %>% distinct(`Check Number`)
# fd %>% distinct(`Check Time`)
# fd %>% distinct(Species)
# fd %>% distinct(Year)
# fd %>% distinct(`Fork Length (mm)`) %>% arrange(`Fork Length (mm)`)



```

## Step 3. Fix defects, 2024

### 2024 Lagoons existence check

```{r}
# fix lagoon names
fd$Lagoon = ifelse(fd$Lagoon=='Aniqaaq','Anigaaq',fd$Lagoon)
fd$Lagoon = ifelse(fd$Lagoon=='Aukulaaq','Aukulak',fd$Lagoon)

# offload non-existent lagoons
fd = fd %>% filter(Lagoon != 'Rabbit Creek' & Lagoon != 'Situkuyok')
# i don't know where rabbit creek or Situkuyok is, no spatial data
# write to a file and filter out of the dataset
#write.csv(fd %>% filter(Lagoon == 'Rabbit Creek'),paste(dir,"L11 Rabbit Creek Fish Data.csv",sep=""))
#write.csv(fd %>% filter(Lagoon == 'Situkuyok'),paste(dir,"L11 Situkuyok Fish Data.csv",sep=""))


# make sure lagoons exist
Lagoons = as.data.frame(fd %>% distinct(Lagoon))

# Rabbit Creek is not a lagoon
for(i in 1:nrow(Lagoons)){
  Lagoon = Lagoons[i,1]
  cat(LagoonExists(connection,Lagoon))
  #cat("INSERT INTO Lagoons(Lagoon)VALUES('",Lagoon,"') -- ",Lagoon," ",Exists,"\n",sep="")
}

# ValidLagoonNames = dbGetQuery(connection,"SELECT DISTINCT Lagoon FROM Lagoons ORDER BY Lagoon")



```

### Fix Sites

```{r}


# Fix Site names
# in order to fix Site we need to move it's current data into SiteType
fd$SiteType = fd$Site

# glimpse(fd)

# Distinct sites
# fd %>% distinct(Site,SiteType) %>% arrange(Site)

# Create site names
fd$Site = paste(toupper(substr(fd$Lagoon,1,3)),"_",toupper(substr(fd$SiteType,1,3)), sep="")

# go back and fix items
fd$Site = ifelse(fd$SiteType=='Inflow',paste(toupper(substr(fd$Lagoon,1,3)),"_IN", sep=""),fd$Site)
fd$Site = ifelse(fd$SiteType=='Marine Edge',paste(toupper(substr(fd$Lagoon,1,3)),"_ME", sep=""),fd$Site)
fd$Site = ifelse(fd$SiteType=='Marine edge',paste(toupper(substr(fd$Lagoon,1,3)),"_ME", sep=""),fd$Site)
fd$Site = ifelse(fd$SiteType=='Inlet Side',paste(toupper(substr(fd$Lagoon,1,3)),"_NS", sep=""),fd$Site)
fd$Lagoon = ifelse(fd$SiteType=='River beaver lodge',"Aukulaaq",fd$Lagoon)
fd$Site = ifelse(fd$SiteType=='River beaver lodge',paste(toupper(substr(fd$Lagoon,1,3)),"_NS", sep=""),fd$Site)

# fd %>% distinct(Site)

# What is River beaver lodge?
#fd %>% filter(SiteType=='River beaver lodge')

```

### 2024 Sites existence check

```{r}

# Check sites exist, if not get insert queries
Sites = fd %>% distinct(Lagoon,Site,Lat,Long)
Sites
cat("BEGIN TRANSACTION -- COMMIT ROLLBACK\n")
for(i in 1:nrow(Sites)){
  Lagoon = Sites[i,'Lagoon']
  Site = Sites[i,'Site']
  Lat = Sites[i,'Lat']
  Lon = Sites[i,'Long']
  cat(SiteExists(connection,Lagoon,Site,Lat,Lon)," -- ",i,"\n",sep="")
}

```

### Fix Sampling Events

```{r}
# Fix Sites
Events = fd %>% distinct(Lagoon,Date,Site) %>% arrange(Site)
Events$RealDate = as.character(Events$Date)

```

### 2024 Sampling Events existence check

```{r}

cat("BEGIN TRANSACTION -- COMMIT ROLLBACK\n")
cat("-- ",nrow(Events)," rows\n")
for(i in 1:nrow(Events)){
  Lagoon = Events[i,'Lagoon']
  Site = Events[i,'Site']
  Date = Events[i,'RealDate']
  cat(SamplingEventExists(connection,Lagoon,Date,Site),"-- ",i,"\n")
}

```



### Fix Gear type
```{r}
# Move gear type to CollectionMethod to preserve gear type
fd$CollectionMethod = fd$`Gear Type`
fd$CollectionMethod = ifelse(fd$`Gear Type`=='offshore gillnet','Gillnet',fd$CollectionMethod)
fd$CollectionMethod = ifelse(fd$`Gear Type`=='onshore gillnet','Gillnet',fd$CollectionMethod)
fd$CollectionMethod = ifelse(fd$`Gear Type`=='beach Seine','Beach Seine',fd$CollectionMethod)
fd$CollectionMethod = ifelse(fd$`Gear Type`=='Onshore gillnet','Gillnet',fd$CollectionMethod)
fd$CollectionMethod = ifelse(fd$`Gear Type`=='Fike','Fyke',fd$CollectionMethod)

fd %>% distinct(`Gear Type`,CollectionMethod)

```

### Fix Set time
```{r}
fd %>% distinct(`Set Time (hours)`)
# These should be in minutes
fd$SetTime_Min = as.integer(fd$`Set Time (hours)`*60)

fd %>% distinct(`Set Time (hours)`,SetTime_Min)


```


### Fix Check Time
```{r}
fd %>% distinct(`Check Time`)
fd$CheckTime= ifelse(!is.na(fd$`Check Time`), paste(fd$`Check Time`,":00",sep=""),NA)

fd %>% distinct(CheckTime)

```
### Fix Species
```{r}
fd %>% distinct(Species)
fd$CommonName = fd$Species

fd$CommonName = ifelse(fd$Species=='No fish',NA,fd$CommonName)
fd$CommonName = ifelse(fd$Species=='No Fish',NA,fd$CommonName)
fd %>% distinct(CommonName)
```

### Fix Fork Length
```{r}

# Fork length is character because of censored data. Some values have a '+', for example
fd$ForkLength = ifelse(is.na(as.numeric(fd$`Fork Length (mm)`)), NA, as.numeric(fd$`Fork Length (mm)`))
# Move text fork lengths into a new column
fd$ForkLengthText = ifelse(is.na(as.numeric(fd$`Fork Length (mm)`)), fd$`Fork Length (mm)`, NA)
# Flag the text fork length data
fd$ForkLengthFlag = ifelse(!is.na(fd$ForkLengthText),6,NA)

# Get the numeric value of fork length text and load it into ForkLength
fd$ForkLength = ifelse(
  is.na(as.numeric(fd$`Fork Length (mm)`)),
  gsub("[^0-9]", "", fd$ForkLengthText)
  ,fd$ForkLength
)
#fd %>% filter(ForkLengthFlag == 6) %>% mutate(ForkLength = gsub("[^0-9]", "", ForkLengthText)) %>% select(ForkLength,ForkLengthText,ForkLengthFlag) 

fd$ForkLength = ifelse(fd$ForkLength == 150300,225,fd$ForkLength)

# check work
#fd %>% select(`Fork Length (mm)`,ForkLength,ForkLengthText,ForkLengthFlag) %>% arrange(ForkLength)
fd %>% distinct(ForkLength,`Fork Length (mm)`,ForkLengthText,ForkLengthFlag) %>% arrange(ForkLengthText)

```

### Fix Counts

```{r}
# Move the text counts to a new column
fd$Count_Text  = NULL

# Fix by moving the text values into a new column
fd$CountText = ifelse(is.na(as.numeric(fd$Count)), fd$Count, NA)

# Flag the text count data
fd$CountFlag = ifelse(!is.na(fd$CountText),7,NA)

# Get the numeric value of count and load it into Count
fd$Count = ifelse(
  is.na(as.numeric(fd$Count)),
  gsub("[^0-9]", "", fd$CountText)
  ,fd$Count
)

# check work
fd %>% distinct(CountText,Count,CountFlag) %>% arrange(CountText,Count)
fd %>% select(Lagoon,Site,Date,CommonName,Count,CountText,CountFlag) %>% arrange(Count,CountText,CountFlag,Lagoon,Site,Date,CommonName)

```
### Fix Lat Lon

```{r}
fd %>% distinct(Lat) %>% arrange(Lat)
fd %>% distinct(Long) %>% arrange(Long)
```



## Step 4. Join the fish scientific names
```{r}
fb = dbGetQuery(connection,"SELECT * from fishbase")
fb$CommonNameUpper = toupper(fb$FBname)
fd$CommonNameUpper = toupper(fd$CommonName)
x = as.data.frame(left_join(fd,fb,by='CommonNameUpper',multiple='first'))
# glimpse(x)

### Fix Species
# x %>% distinct(Genus,Species.y) %>% arrange(Genus,Species.y)

```


## Step 5. Write the fish deliverable

```{r}


deliverable = x %>% select(Date
,Lagoon
,Site
,Family
,Genus
,`Species.y`
,CommonName
,CollectionMethod
,`Check Number` 
,CheckTime
,SetTime_Min
,Count,CountText,CountFlag
,ForkLength,ForkLengthText,ForkLengthFlag
,Lat
,Long
,`Species.x`
,Notes,SiteType)

# deliverable = sqldf("SELECT Date
# ,Lagoon
# ,Site
# ,Family
# ,Genus
# ,[Species.y] as Species
# ,CommonName
# ,CollectionMethod
# ,[Check Number] as CheckNumber
# ,CheckTime
# ,SetTime_Min
# ,Count,CountText,CountFlag
# ,ForkLength as ForkLength_mm,ForkLengthText,ForkLengthFlag as ForkLengthFlag
# ,Lat as Latitude
# ,Long as Longitude
# ,[Species.x] as CommonNameRecorded
# ,Notes as Comments,SiteType as Site_Recorded
# FROM x 
# ORDER BY Date
# ,Lagoon
# ,Site,CollectionMethod,CheckNumber")

write.csv(deliverable,paste(dir,"L11 2024 Lagoons Fish Data.csv",sep=""),row.names = FALSE,na="")

```


# 2021 Deliverable L10-Discrete water quality data processing


## Step 1. Load the worksheet

```{r}
# Get the worksheet names
#excel_sheets(workbook)

# Load the fish data
worksheet = "Water Data 2021"
wd = readxl::read_excel(workbook,worksheet)

```

## Step 2. Data quality checks

```{r}

# Use these tools to check that data types match expected attributes, statistics such as min/max, etc. are as expected
glimpse(wd)

# Use these tools to check that data types match expected attributes, statistics such as min/max, etc. are as expected
wd$Date = as.Date(wd$Date)

# Generate dplyr distinct values statements
for(ColumnName in colnames(wd)){
  cat("wd %>% distinct(`",ColumnName,"`) %>% arrange(`",ColumnName,"`)\n",sep="")
}

wd %>% distinct(`Lagoon`) %>% arrange(`Lagoon`)
wd %>% distinct(`Date`) %>% arrange(`Date`)
wd %>% distinct(`Site`) %>% arrange(`Site`)
wd %>% distinct(`SiteType`) %>% arrange(`SiteType`)
wd %>% distinct(`Temp_C`) %>% arrange(`Temp_C`)
wd %>% distinct(`Conductivity (µS/cm)`) %>% arrange(`Conductivity (µS/cm)`)
wd %>% distinct(`Salinity (ppt)`) %>% arrange(`Salinity (ppt)`)
wd %>% distinct(`ODO%`) %>% arrange(`ODO%`)
wd %>% distinct(`ODO mg/L`) %>% arrange(`ODO mg/L`)
wd %>% distinct(`pH`) %>% arrange(`pH`)
wd %>% distinct(`Chlorophyl_RFU`) %>% arrange(`Chlorophyl_RFU`)
wd %>% distinct(`BGA RFU`) %>% arrange(`BGA RFU`)
wd %>% distinct(`Turbidity (FNU)`) %>% arrange(`Turbidity (FNU)`)
wd %>% distinct(`NOTES`) %>% arrange(`NOTES`)
wd %>% distinct(`ExcelDate`) %>% arrange(`ExcelDate`)

```

## Step 3. Fix defects, 2024

### 2024 Lagoons existence check

```{r}
# fix lagoon names
# wd$Lagoon = ifelse(wd$Lagoon=='Aniqaaq','Anigaaq',wd$Lagoon)
# wd$Lagoon = ifelse(wd$Lagoon=='Aukulaaq','Aukulak',wd$Lagoon)


# make sure lagoons exist
Lagoons = as.data.frame(wd %>% distinct(Lagoon))

# Rabbit Creek is not a lagoon
for(i in 1:nrow(Lagoons)){
  Lagoon = Lagoons[i,1]
  cat(LagoonExists(connection,Lagoon))
  #cat("INSERT INTO Lagoons(Lagoon)VALUES('",Lagoon,"') -- ",Lagoon," ",Exists,"\n",sep="")
}

# ValidLagoonNames = dbGetQuery(connection,"SELECT DISTINCT Lagoon FROM Lagoons ORDER BY Lagoon")



```

### Fix Sites

```{r}


# Fix Site names
# in order to fix Site we need to move it's current data into SiteType
#wd$SiteType = wd$Site

# glimpse(wd)

# Distinct sites
# wd %>% distinct(Site,SiteType) %>% arrange(Site)

# Create site names
#wd$Site = paste(toupper(substr(wd$Lagoon,1,3)),"_",toupper(substr(wd$SiteType,1,3)), sep="")

# go back and fix items
# wd$Site = ifelse(wd$SiteType=='Inflow',paste(toupper(substr(wd$Lagoon,1,3)),"_IN", sep=""),wd$Site)
# wd$Site = ifelse(wd$SiteType=='Marine Edge',paste(toupper(substr(wd$Lagoon,1,3)),"_ME", sep=""),wd$Site)
# wd$Site = ifelse(wd$SiteType=='Marine edge',paste(toupper(substr(wd$Lagoon,1,3)),"_ME", sep=""),wd$Site)
# wd$Site = ifelse(wd$SiteType=='Inlet Side',paste(toupper(substr(wd$Lagoon,1,3)),"_NS", sep=""),wd$Site)
# wd$Lagoon = ifelse(wd$SiteType=='River beaver lodge',"Aukulaaq",wd$Lagoon)
# wd$Site = ifelse(wd$SiteType=='River beaver lodge',paste(toupper(substr(wd$Lagoon,1,3)),"_NS", sep=""),wd$Site)

# wd %>% distinct(Site)

# What is River beaver lodge?
#wd %>% filter(SiteType=='River beaver lodge')

```

### 2024 Sites existence check

```{r}
wd$Lat = NA
wd$Long = NA
# Check sites exist, if not get insert queries
Sites = wd %>% distinct(Lagoon,Site,Lat,Long)
Sites
cat("BEGIN TRANSACTION -- COMMIT ROLLBACK\n")
for(i in 1:nrow(Sites)){
  Lagoon = Sites[i,'Lagoon']
  Site = Sites[i,'Site']
  Lat = Sites[i,'Lat']
  Lon = Sites[i,'Long']
  cat(SiteExists(connection,Lagoon,Site,Lat,Lon)," -- ",i,"\n",sep="")
}

```



### 2024 Sampling Events existence check

```{r}
Events = wd %>% distinct(Lagoon,Date,Site) %>% arrange(Site)
Events$RealDate = as.character(Events$Date)

cat("BEGIN TRANSACTION -- COMMIT ROLLBACK\n")
cat("-- ",nrow(Events)," rows\n")
for(i in 1:nrow(Events)){
  Lagoon = Events[i,'Lagoon']
  Site = Events[i,'Site']
  Date = Events[i,'RealDate']
  cat(SamplingEventExists(connection,Lagoon,Date,Site),"-- ",i,"\n")
}

```

## Step 5. Write the water quality deliverable

```{r}


deliverable = x %>% select(Date
,Lagoon
,Site
,Family
,Genus
,`Species.y`
,CommonName
,CollectionMethod
,`Check Number` 
,CheckTime
,SetTime_Min
,Count,CountText,CountFlag
,ForkLength,ForkLengthText,ForkLengthFlag
,Lat
,Long
,`Species.x`
,Notes,SiteType)

# deliverable = sqldf("SELECT Date
# ,Lagoon
# ,Site
# ,Family
# ,Genus
# ,[Species.y] as Species
# ,CommonName
# ,CollectionMethod
# ,[Check Number] as CheckNumber
# ,CheckTime
# ,SetTime_Min
# ,Count,CountText,CountFlag
# ,ForkLength as ForkLength_mm,ForkLengthText,ForkLengthFlag as ForkLengthFlag
# ,Lat as Latitude
# ,Long as Longitude
# ,[Species.x] as CommonNameRecorded
# ,Notes as Comments,SiteType as Site_Recorded
# FROM x 
# ORDER BY Date
# ,Lagoon
# ,Site,CollectionMethod,CheckNumber")

write.csv(deliverable,paste(dir,"L11 2024 Lagoons Fish Data.csv",sep=""),row.names = FALSE,na="")

```




