---
title: "2024 Lagoons Data ETL"
author: "SDMiller"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
#library(RODBC)
# library(knitr)
# library(ggplot2)
# library(sp)
# library(sf)
# library(sqldf)
library(tidyverse)
library(readxl)
# library(odbc)
# library(leaflet)

connection = dbConnect(odbc(),Driver = "Sql Server",Server = "inpyugamsvm01\\nuna_dev", Database = "ARCN_Lagoons")
dir = r'(O:\Monitoring\Vital Signs\Lagoon Communities and Ecosystems\Data\2024 Lagoons Sampling\Deliverables processing/)'
dir = r'(C:\Work\VitalSigns\ARCN Lagoons\Data\2024 Lagoons Sampling\Deliverables processing/)' # copied local for speed
filename = r'(2021-2024 WCS CAKR Coastal Lagoons Data Received 2024-07-31 From KFraley.xlsx)'
workbook = paste(dir,filename,sep="")

```

# Deliverable L11-Fish data processing

This section contains processing steps to get the fish data (Deliverable L11) processed for ingestion into the master lagoons monitoring database.

## Step 1. Load the worksheet

```{r}
# Get the worksheet names
#excel_sheets(workbook)

# Load the fish data
worksheet = "Fish Data 2024"
fd = readxl::read_excel(workbook,worksheet)

```

## Step 2. Check data types

```{r}

# Use these tools to check that data types match expected attributes, statistics such as min/max, etc. are as expected
# glimpse(fd)

# view(fd)


```

## Step 3. Fix defects

```{r}
# Dates are not real dates
fd$Date = as.Date(fd$Date)

```


# Step 1. Ensure the Lagoons in the Water Quality worksheet validate with standard Lagoon names in the database

Lagoons must match existing lagoons, as spelled in the database Lagoons table.

```{r}

# First, make sure the lagoons exist in the lagoons table
x = as.data.frame(wq %>% distinct(Lagoon))
for(i in 1:nrow(x)){
  y=x[i,1]
  sql = paste("SELECT * FROM Lagoons WHERE Lagoon='",y,"' -- INSERT INTO Lagoons(Lagoon)VALUES('",y,"')\n",sep="")
  cat(sql)
}

```

# Insert the Sites if they don't exist

```{r}


# Check that the field sites exist in the Sites database table. If the field sites do not exist in the database then those records will be rejected by the database.
# This code writes to standard output a set of SELECT queries that can be executed against the database. If any SELECT queries return zero results, then you must INSERT the site using the accompanying INSERT query.
x = as.data.frame(wq %>% distinct(Lagoon,Location))
SourceFile='L10 2024 Lagoons Discrete Water Quality Data.csv'
for(i in 1:nrow(x)){
  Lagoon=x[i,1]
  Site = x[i,2]
  sql = paste("SELECT * FROM Sites WHERE Lagoon='",Lagoon,"' And  Site='",Site,"'\n -- INSERT INTO Sites(Lagoon,Site,SourceFile)VALUES('",Lagoon,"','",Site,"','",SourceFile,"')\n",sep="")
  cat(sql)
}
# The sites should all exist now

```

# Insert sampling events

```{r}
# Next insert the sampling events

x = as.data.frame(wq %>% distinct(Lagoon,Location,Date))
SourceFile='L10 2024 Lagoons Discrete Water Quality Data.csv'
Output = 'BEGIN TRANSACTION -- COMMIT ROLLBACK\n'
for(i in 1:nrow(x)){
  Lagoon=x[i,1]
  Site = x[i,2]
  Date=x[i,3]
  Sql = paste("INSERT INTO SamplingEvents(Lagoon,Site,Date,SourceFile)VALUES('",Lagoon,"','",Site,"','",Date,"','",SourceFile,"')\n",sep="")
  Output = paste(Output,Sql,sep="")
}
write(Output,paste(dir,"INSERT SamplingEvents.sql",sep=""))

```

# Insert discrete water quality data

```{r}
# Do a little quality control
# glimpse(wq)
# # ODO mg/L is all NULLs
# wq %>% distinct(`ODO mg/L`)
# # Notes is all NULLs
# wq %>% distinct(NOTES)
# library(summarytools)
# descr(wq)
# str(wq)


wq %>% mutate(DateAK = with_tz(Date, tzone = "America/Anchorage"))
wq$SPC = wq$`SPC (µS/cm)`
wq$SPC
CSVFile='L10 2024 Lagoons Discrete Water Quality Data.csv'
SqlFile=paste(CSVFile,".sql",sep="")
Output = 'BEGIN TRANSACTION -- COMMIT ROLLBACK\n'
for(i in 1:nrow(wq)){
  Lagoon = ifelse(is.na(wq[i, 'Lagoon']),'NULL', wq[i, 'Lagoon'])
  Date = ifelse(is.na(wq[i, 'DateAK']),'NULL', wq[i, 'DateAK'])
  Location = ifelse(is.na(wq[i, 'Location']),'NULL', wq[i, 'Location'])
  TEMP = ifelse(is.na(wq[i, 'TEMP (Celsius)']),'NULL', wq[i, 'TEMP (Celsius)'])
  SPC = wq[i, 'SPC']
  Salinity = ifelse(is.na(wq[i, 'Salinity (ppt)']),'NULL', wq[i, 'Salinity (ppt)'])
  ODOPct = ifelse(is.na(wq[i, 'ODO%']),'NULL', wq[i, 'ODO%'])
  ODOmgL = ifelse(is.na(wq[i, 'ODO mg/L']),'NULL', wq[i, 'ODO mg/L'])
  pH = ifelse(is.na(wq[i, 'pH']),'NULL', wq[i, 'pH'])
  ChlorophylRFU = ifelse(is.na(wq[i, 'Chlorophyl RFU']),'NULL', wq[i, 'Chlorophyl RFU'])
  BGARFU = ifelse(is.na(wq[i, 'BGA RFU']),'NULL', wq[i, 'BGA RFU'])
  TurbidityFNU = ifelse(is.na(wq[i, 'Turbidity (FNU)']),'NULL', wq[i, 'Turbidity (FNU)'])
  NOTES = ifelse(is.na(wq[i, 'NOTES']),'NULL', wq[i, 'NOTES'])

    Sql = paste("INSERT INTO WaterQualityDiscrete(
Year
,Lagoon
,Site
,Date
,LocationDescription
,[Temperature (C)]
,[Conductivity (uS/cm)]
,[Salinity (ppt)]
,[Dissolved Oxygen (% sat)]
,[Dissolved Oxygen (mg/l)]
,[pH]
,[Chlorophyll (RFU)]
,[Blue-green Algae  (RFU)]
,[Turbidity (FNU)]
,SourceFile)VALUES(
'",Lagoon,"'
,'",Site,"'
,'",Date,"'
,'",Location,"'
,",TEMP,"
,",SPC,"
,",Salinity,"
,",ODOPct,"
,",ODOmgL,"
,",pH,"
,",ChlorophylRFU,"
,",BGARFU,"
,",TurbidityFNU,"
,'",SourceFile,"')\n",sep="")
  Output = paste(Output,Sql,sep="")
}
write(Output,paste(dir,SqlFile,sep=""))
write.table(wq,paste(dir,CSVFile,sep=""),row.names = FALSE,na = "NULL",quote = FALSE,sep = ",")
writexl::write_xlsx(wq,paste(dir,CSVFile,".xlsx",sep=""))
# pre-print columnname definitions
# for (colname in colnames(wq)) {
#   
#   cat(gsub(')','',gsub(' ','',colname))," = wq[i,'",colname,"']","\n",sep="")
# }

```



# WQ EA INSERT

The below was a test to see how hard it would be to extract and store the WQD data as entity/value pairs. It proved difficult but could be useful for parameters that fall outsite the 'standard' ones described by the protocol.

```{r}
wq = readxl::read_excel(workbook,worksheet)
wq = as.data.frame(wq)
# glimpse(wq)
# Next insert the data

# Output = 'BEGIN TRANSACTION -- COMMIT ROLLBACK\n'
# 
# GetInsertQuery = function(ColumnName,DesiredColumnName,Units){
#   for(i in 1:nrow(wq)){
#     Lagoon = wq[i,'Lagoon']
#     Date  = wq[i,'Date']
#     Site  = wq[i,'Location']
#     Entity = ColumnName
#     Value = wq[i,ColumnName]
#     
#     Sql = paste("INSERT INTO [dbo].[WaterQualityDiscreteEA]
#            ([Lagoon]
#            ,[Site]
#            ,[Date]
#            ,[Entity]
#            ,[Value]
#            ,[SampleDepth_m]
#            ,[Units]
#            ,SourceFile)
#      VALUES
#            ('",Lagoon,"'
#            ,'",Site,"'
#            ,'",Date,"'
#            ,'",DesiredColumnName,"'
#            ,",wq[i,ColumnName],"
#            ,NULL
#            ,'",Units,"'
#            ,'",SourceFile,"')\n\n",sep="")
#     Output = paste(Output,Sql,sep="")
#   }
#   return(Output)
# }
# ColumnName = 'TEMP (Celsius)'
# DesiredColumnName = 'Temperature'
# Units = 'C'
# Queries = GetInsertQuery(ColumnName,DesiredColumnName,Units)
# OutputFilename = paste(dir,"INSERT ",DesiredColumnName,".sql",sep="")
# write(Queries,OutputFilename)

# pre-print columnname definitions
# for (colname in colnames(wq)) {
#   cat(colname," = wq[i,'",colname,"']","\n")
# }


```

