---
title: "ARCN Data Series Report: Coastal Lagoons Monitoring in Bering Land Bridge National Preserve and Cape Krusenstern National Monument"
author: "SDMiller"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

# Introduction

The National Park Service (NPS) has monitored the ecology of the coast lagoons in Bering Land Bridge National Preserve and Cape Krusenstern National Monument since 2003 as part of the Arctic Inventory and Monitoring Network (ARCN) initiative (Lawler et al., 2009). This report is intended to provide a visual representation of the data for quality control purposes. Each of the main water quality and fish parameters is summarized by lagoon, year and site type in both tabular and visual formats, allowing the reader to see and quickly identify data quality problems for rectification and/or documentation.

# Methods

Lagoon ecosystem monitoring methods are described in Jones et al., (2018) and Jones (2017).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Load libraries
library(sqldf)
library(tidyverse)
library(readxl)
library(odbc)
library(leaflet)
library(leaflet.extras)
library(leafem)

# Database connection
Connection = dbConnect(odbc(),Driver = "Sql Server",Server = "inpyugamsvm01\\nuna", Database = "ARCN_Lagoons")

# Table and figure counters
TableCounter = 1
FigureCounter = 1

```

```{r,label="Database queries",echo=FALSE,warning=FALSE,message=FALSE}
# Get the discrete water quality data
Sql = "SELECT * FROM WaterQualityDiscrete"
wqd = dbGetQuery(Connection,Sql)

# Make dates actual dates
wqd$Date = as.Date(wqd$Date)

# If SiteType is null convert to Non-standard
wqd$SiteType = ifelse(is.na(wqd$SiteType)==TRUE,'Non-standard',wqd$SiteType)

# Get the discrete water quality data
Sql = "SELECT [Lagoon]
      ,[Park]
      ,[OceanConnectedness]
      ,Convert(Varchar(4000),[Description]) as Description
      ,[Latitude]
      ,[Longitude]
      --,[PolygonFeature]
      --,[RecordInsertedDate]
      --,[RecordInsertedBy]
  FROM [ARCN_Lagoons].[dbo].[Lagoons]
  WHERE NOT Lagoon in ('Ocean')
  ORDER BY Lagoon"
Lagoons = dbGetQuery(Connection,Sql)

# Summarize the number of sampling days by year by lagoon
WaterQualityDiscreteSamplingHistory = wqd %>% group_by(Lagoon,Year) %>% summarise(`Sampling days`=n_distinct(Date),Earliest=min(Date),Latest=max(Date)) %>% filter(`Sampling days` > 0 & Lagoon !='Ocean') %>% arrange(Lagoon,Year)


```

```{r,label='Functions',echo=FALSE,warning=FALSE,message=FALSE}

# Returns a data frame of summary statistics for the discrete water quality parameter Parameter for Lagoon
# LagoonName: The name of a lagoon from the Lagoons database table
# Parameter from the WaterQualityDiscrete database table.
# Returns: Tibble
GetWaterQualityDiscreteParameterSummaryForALagoon = function(LagoonName,Parameter){
    # Build a summary of basic statistics for the lagoon and parameter
    Summary = wqd %>% filter(Lagoon==LagoonName & is.na(!!sym(Parameter))==FALSE) %>% group_by(Date) %>% 
      summarize(
        Mean = mean(!!sym(Parameter),na.rm=TRUE)
        ,Min = min(!!sym(Parameter),na.rm=TRUE)
        ,Max = max(!!sym(Parameter),na.rm=TRUE)
        ,SD = sd(!!sym(Parameter),na.rm=TRUE)
        ,n = n()
        ,.groups='keep'
      ) %>%
      arrange(Date)
  return(Summary)
}
#Test
# GetWaterQualityDiscreteParameterSummaryForALagoon('Kotlik','Conductivity (uS/cm)')
# GetWaterQualityDiscreteParameterSummaryForALagoon('Krusenstern','Conductivity (uS/cm)')


PlotWaterQualityDiscreteParameterSummaryForALagoon = function(LagoonName,Parameter,FigureNumber){
  # Get a table of basic statistics for the lagoon and parameter
  Summary = GetWaterQualityDiscreteParameterSummaryForALagoon(LagoonName,Parameter) 
  
  # Plot the parameter
  Plot = ggplot(Summary) +
    geom_ribbon(aes(x=Date,ymin=Min,ymax=Max,group=1),fill='gray',alpha=0.5) +
    geom_errorbar(aes(x=Date,ymin=Mean-SD,ymax=Mean+SD,group=1),width=0.3) +
    geom_path(aes(x=Date,y=Mean,group = 1),na.rm=TRUE) +
    geom_point(aes(x=Date,y=Mean),fill='black',na.rm=TRUE) +
    labs(
      title=paste(Parameter," in ",LagoonName," lagoon",sep=""),
      caption=paste("Figure ",FigureNumber,". ",Parameter," in ",LagoonName," lagoon. Error bars indicate standard deviation. Gray band shows minimum and maximum daily values.",sep="")
      ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
  
  return(Plot)
}
# Test
 PlotWaterQualityDiscreteParameterSummaryForALagoon('Kotlik','Salinity (ppt)',3)
# PlotWaterQualityDiscreteParameterSummaryForALagoon('Krusenstern','Conductivity (uS/cm)',3)

 
 
 
 # Returns a data frame of summary statistics for the discrete water quality parameter Parameter for Lagoon by site type (random, marine edge, inflow, etc.)
# LagoonName: The name of a lagoon from the Lagoons database table
# Parameter from the WaterQualityDiscrete database table.
# Returns: Tibble
GetWaterQualityDiscreteParameterSummaryForALagoonAndSiteType = function(LagoonName,Parameter){
  Summary = wqd %>% 
    mutate(SiteType=toupper(SiteType)) %>%
    select(Lagoon,SiteType,Date,!!sym(Parameter)) %>%
    filter(Lagoon==LagoonName & is.na(!!sym(Parameter))==FALSE) %>% 
    mutate(SiteType = ifelse(is.na(SiteType),toupper('Non-standard'),SiteType)) %>%
    group_by(SiteType,Date) %>%
    summarize(
      Mean = mean(!!sym(Parameter),na.rm=TRUE)
      ,Min = min(!!sym(Parameter),na.rm=TRUE)
      ,Max = max(!!sym(Parameter),na.rm=TRUE)
      ,SD = sd(!!sym(Parameter),na.rm=TRUE)
      ,n = n()
      ,.groups='keep'
    ) %>%
    arrange(Lagoon,SiteType,Date)
  
  return(Summary)
}
#Test
# GetWaterQualityDiscreteParameterSummaryForALagoonAndSiteType('Aukulak','Conductivity (uS/cm)')
# GetWaterQualityDiscreteParameterSummaryForALagoon('Krusenstern','Conductivity (uS/cm)')


 
PlotWaterQualityDiscreteParameterSummaryForALagoonAndSiteType = function(LagoonName,Parameter,FigureNumber){
  # Get a table of basic statistics for the lagoon and parameter
  Summary = GetWaterQualityDiscreteParameterSummaryForALagoonAndSiteType(LagoonName,Parameter) 
  
  # Plot the parameter
  Plot = ggplot(Summary) +
    geom_ribbon(aes(x=Date,ymin=Min,ymax=Max,group=1),fill='gray',alpha=0.5) +
    geom_errorbar(aes(x=Date,ymin=Mean-SD,ymax=Mean+SD),width=0.5) +
    geom_path(aes(x=Date,y=Mean,linetype=SiteType),na.rm=TRUE) +
    geom_point(aes(x=Date,y=Mean),fill='black',na.rm=TRUE) +
    labs(
      title=paste(Parameter," in ",LagoonName," lagoon",sep=""),
      caption=paste("Figure ",FigureNumber,". ",Parameter," in ",LagoonName," lagoon. Error bars indicate standard deviation. Gray band shows minimum and maximum daily values.",sep="")
      ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 
  
  return(Plot)
}
# Test
# PlotWaterQualityDiscreteParameterSummaryForALagoonAndSiteType('Aukulak','Salinity (ppt)',3)
# PlotWaterQualityDiscreteParameterSummaryForALagoonAndSiteType('Aukulak','Conductivity (uS/cm)',3)
#  
 
 
 
 
# Get a column plot of sampling days by year by lagoon
GetWaterQualityDiscreteSamplingHistoryPlot = function(LagoonName,FigureNumber){
  Title = paste("Discrete water quality sampling history in ",LagoonName," lagoon.",sep="")
  # Plot the number of sampling days by year by lagoon
  Plot = ggplot(WaterQualityDiscreteSamplingHistory %>% filter(Lagoon == LagoonName)) +
    geom_col(aes(x = Year, y = `Sampling days`),width=0.3) +
    labs(title=Title,caption=paste("Figure ",FigureNumber,". ",Title,".",sep=""),hjust=0) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    scale_x_continuous(breaks = seq(min(WaterQualityDiscreteSamplingHistory$Year), max(WaterQualityDiscreteSamplingHistory$Year), by = 1))
  return(Plot)
}
# Test: 
# GetWaterQualityDiscreteSamplingHistoryPlot('Atiligauraq',3)
# GetWaterQualityDiscreteSamplingHistoryPlot('Kotlik',3)



# Returns a leaflet map of LagoonName
# Lagoon: Lagoon to map
# Returns: Leaflet map
GetLagoonMap = function(LagoonName){
  Map = leaflet(Lagoons %>% filter(Lagoon==LagoonName)) %>%
    # Points
    addCircleMarkers(lng = ~Longitude, lat = ~Latitude,label=~Lagoon,color='red',fillColor = 'red',stroke=FALSE,fillOpacity = 0.5,
    labelOptions = labelOptions(noHide = TRUE, fill = FALSE, offset = c(5,2), textsize = "18px", textOnly = TRUE)
    ) %>%
    
    # Background map (https://leaflet-extras.github.io/leaflet-providers/preview/)
    #addTiles()  
    #addProviderTiles("Esri.WorldImagery") # Imagery, no contextual labels.
    #addProviderTiles("Stadia.StamenTerrain") # Nice terrain and colors but no contextual labels
    #addProviderTiles("Esri.WorldTopoMap") # Very light, feature labels
    #addProviderTiles("Esri.NatGeoWorldMap") # Nice maps, better as you zoom in. Labels.
    addProviderTiles("USGS.USTopo")
    
    return(Map)
}
# Test
# GetLagoonMap('Aukulak')



# This function bins water quality discrete data, counts the members in each bin, and then plots the counts as a histogram
# LagoonName: Lagoon
# Parameter: Water quality parameter. Must match a column in the wqd dataset
# Returns: Histogram plot of counts by values for Parameter
GetHistogramOfWaterQualityDiscreteParameter = function(LagoonName,Parameter,FigureCounter){
  #LagoonName='ocean'
  #Parameter='Temperature (C)'
  
  # Bin the water quality data into 20 bins by Parameter 
  Data =  wqd %>% filter(Lagoon==LagoonName & is.na(!!sym(Parameter))==FALSE) %>% select(Lagoon,!!sym(Parameter))
  if(nrow(Data)>0){
    
    # Bin the data into counts
    Summary = Data %>% mutate(Bin = cut(!!sym(Parameter), breaks = 20)) %>% count(Bin)

    # Plot the counts of each bin
    Title = paste("Histogram of ",Parameter," values: ",LagoonName,sep="")
    ggplot(Summary, aes(x = Bin, y = n)) +
        geom_col() +
        labs(title = "Binned Dissolved Oxygen Counts", x = "Dissolved Oxygen (mg/L) Bins", y = "Count") +
        labs(title=Title,caption=paste("Figure ",FigureCounter,". ",Title,".",sep=""),hjust=0) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate labels for readability
  }
}
#Test
# GetHistogramOfWaterQualityDiscreteParameter('Aukulak','Temperature (C)',3)


```

# Study Sites

NPS monitors fish and water quality in the coastal lagoons of northwester Alaska (Figure `r FigureCounter`). Within each lagoon samples are taken from a variety of standard sites (inflow, outflow, marine edge, terrestrial edge and random) as well as opportunistic non-standard sites in areas of interest like main channels, the centers of the lagoons, the ocean external to the lagoons, etc.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Map the points
leaflet(Lagoons) %>%
    # Points
    addCircleMarkers(lng = ~Longitude, lat = ~Latitude,label=~Lagoon,color='red',fillColor = 'red',stroke=FALSE,fillOpacity = 0.5,
    labelOptions = labelOptions(noHide = TRUE, fill = FALSE, offset = c(5,2), textsize = "18px", textOnly = TRUE)
    ) %>%
    
    # Background map. For examples see https://leaflet-extras.github.io/leaflet-providers/preview/
    #addTiles()  
    #addProviderTiles("Esri.WorldImagery") %>% # Imagery, no contextual labels.
    #addProviderTiles("Stadia.StamenTerrain") %>% # Nice terrain and colors but no contextual labels
    #addProviderTiles("Esri.WorldTopoMap") %>%  # Very light, feature labels
    #addProviderTiles("Esri.NatGeoWorldMap") %>%  # Nice maps, better as you zoom in. Labels.
    addProviderTiles("USGS.USTopo") %>%
    addMouseCoordinates()  # Display mouse coordinates (lat/lon)
    
```

# Discrete Water Quality Sampling Results

## Sampling History

```{r,echo=FALSE,warning=FALSE,message=FALSE,fig.height=10,fig.width=10}


# Plot the number of sampling days by year by lagoon
ggplot(WaterQualityDiscreteSamplingHistory) +
  geom_col(aes(x = Year, y = `Sampling days`)) +
  geom_text(aes(x=min(Year),y=max(`Sampling days`) + 5,label = Lagoon),size=3,hjust=0) +
  facet_grid(vars(Lagoon)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),  # Rotates x-axis labels
    #strip.text.x = element_text(angle = 0)  # Keeps facet grid labels horizontal
    strip.text = element_blank()  # Removes facet grid labels
  ) +
  scale_x_continuous(breaks = seq(min(WaterQualityDiscreteSamplingHistory$Year), max(WaterQualityDiscreteSamplingHistory$Year), by = 1))  # Sets breaks incrementing by 1



```

## Water Quality Parameters By Lagoon

```{r,results='asis',echo=FALSE,warning=FALSE,message=FALSE}

# This function prints out a map of the lagoon, a sampling history plot and a summary table and plot of each main chemical or physical water quality parameter.
# LagoonName: Name of the lagoon to profile.
# Returns: Markdown report section.
ProfileALagoon = function(LagoonName){
  
    # Print the lagoon name as markdown
  cat("  \n\n",sep="")
  cat("### ",LagoonName,"  \n\n",sep="")
  cat("  \n\n",sep="")
  cat("This section of the report characterizes ",as.character(LagoonName)," Lagoon and summarizes our sampling history and results of discrete physical and chemical data. \n\n",sep="")
  cat("  \n\n",sep="")
  
  # Print a map of the lagoon
  cat("  \n\n",sep="")
  print(GetLagoonMap(LagoonName)) # GetLagoonMap('Kotlik')
  cat("Figure ",FigureCounter,". Map of ",LagoonName," Lagoon.  \n\n",sep="")
  cat("  \n\n",sep="")
  FigureCounter=FigureCounter+1
  
  # Print a column chart of number of sampling days per year for the lagoon - a visual sampling history
  cat("  \n\n",sep="")
  print(GetWaterQualityDiscreteSamplingHistoryPlot(LagoonName,FigureCounter))
  cat("  \n\n",sep="")
  FigureCounter=FigureCounter +1 
  
  # Make a list of discrete water quality parameters to describe
  Parameters = c("Conductivity (uS/cm)","Salinity (ppt)" ,"Salinity (PSU)","Dissolved Oxygen (% sat)","Dissolved Oxygen (mg/l)","pH","Turbidity (FNU)","Turbidity (NTU)","Turbidity (mg/L)","Chlorophyll (ug/L)")
  
  # Loop through the parameters and output results as markdown, one by one
  for(Parameter in Parameters){
    
    # Parameter name
    cat("  \n\n",sep="")
    cat("#### ",Parameter,"  \n\n",sep="")
    cat("  \n\n",sep="")
    
    # Plot the parameter
    #print(PlotWaterQualityDiscreteParameterSummaryForALagoon(LagoonName,Parameter,FigureCounter))
    cat("  \n\n",sep="")
    print(PlotWaterQualityDiscreteParameterSummaryForALagoonAndSiteType(LagoonName,Parameter,FigureCounter))
    cat("  \n\n",sep="")
    FigureCounter = FigureCounter + 1
    
    # Give a summary table
    TableCaption = paste("Table ",TableCounter,". Summary of ",Parameter," in ",LagoonName," lagoon. \n\n",sep="")
    cat("  \n\n",sep="")
    print(knitr::kable(GetWaterQualityDiscreteParameterSummaryForALagoonAndSiteType(LagoonName,Parameter),caption=TableCaption))
    cat("  \n\n",sep="")
    TableCounter = TableCounter + 1
    
    # Histogram the results to look for outliers
    cat("  \n\n",sep="")
    print(GetHistogramOfWaterQualityDiscreteParameter(LagoonName,Parameter,FigureCounter))
    cat("  \n\n",sep="")
    FigureCounter = FigureCounter + 1
  }
}
# TEST: ProfileALagoon('Aukulak')

# Profile the lagoons one by one
for(Lagoon in Lagoons$Lagoon){
  LagoonName = Lagoon
  ProfileALagoon(as.character(LagoonName))
}

```

# Continuous Water Quality Sampling Results

```{r,echo=FALSE,echo=FALSE,warning=FALSE,message=FALSE}
# Get the continuous water quality data
Sql = "SELECT * FROM WaterQualityContinuous"
wqc = dbGetQuery(Connection,Sql)
wqc$Date = as.Date(wqc$DateTime)
#  glimpse(wqc)

```

```{r,results='asis',echo=FALSE,warning=FALSE,message=FALSE}
# This function prints out a map of the lagoon, a sampling history plot and a summary table and plot of each main chemical or physical water quality parameter.
# LagoonName: Name of the lagoon to profile.
# Returns: Markdown report section.
ProfileContinuousWaterQuality = function(LagoonName){
  # Print the lagoon name as markdown
  cat("## ",as.character(LagoonName)," \n\n",sep="")
  cat("This section of the report characterizes the continuous water quality data for ",as.character(LagoonName)," Lagoon. \n\n",sep="") 

    # Make a list of discrete water quality parameters to describe
  Parameters = c("Turbidity (FNU)"
,"TSS (mg/L)"
,"Chlorophyll (RFU)"
,"Chlorophyll (ug/L)"
,"Blue-green algae (RFU)"
,"Blue-green algae (ug/L)"
,"Temperature (C)"
,"Conductivity (uS/cm)"
,"Salinity (psu)"
,"Salinity (ppt)"
,"nLF Cond (uS/cm)"
,"TDS (mg/L)"
,"ODO (% sat)"
,"ODO (mg/L)"
,"pH"
,"Oxidative reduction potential (mV)"
,"[Oxidative reduction potential raw mV"
,"Pressure (psi)"
,"Depth (m)")
  
  # Loop through the parameters and output results as markdown, one by one
  for(Parameter in Parameters){
    # Parameter name
    cat("### ",Parameter," \n\n",sep="")
    # Summarize the data by lagoon and date
    Summary = wqc %>% filter(Lagoon==LagoonName & is.na(!!sym(Parameter))==FALSE) %>% group_by(Date) %>% summarize(
      Mean=mean(!!sym(Parameter), na.rm = TRUE)
      ,Min=min(!!sym(Parameter), na.rm = TRUE)
      ,Max=max(!!sym(Parameter), na.rm = TRUE)
      ,`S.D.`=sd(!!sym(Parameter), na.rm = TRUE)
      ,n=n()
    )
    
    # Plot the Parameter
    Title=paste(LagoonName,": ",Parameter,sep="")
    Caption = paste("Figure ",FigureCounter,". ",Parameter," in ",LagoonName,".",sep="")
    print(ggplot(Summary) +
        geom_point(aes(x=Date,y=Mean)) +
        geom_line(aes(x=Date,y=Mean)) +
        geom_errorbar(aes(x=Date,ymin=Mean-`S.D.`,ymax=Mean+`S.D.`),width=0.3) +
        geom_ribbon(aes(x=Date,ymin=Min,ymax=Max),fill='steelblue',alpha=0.5) +
        labs(title=Title,caption=Caption,hjust=0) +
        scale_x_date(date_labels = "%Y-%m-%d") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 90, hjust = 1))
    )
    FigureCounter = FigureCounter + 1
    
    # Output the summary table
    print(knitr::kable(Summary,caption = paste("Table ",TableCounter,". ",Parameter," in ",LagoonName,sep="")))
    TableCounter = TableCounter+1
    
  }
}
# ProfileContinuousWaterQuality('Anigaaq')

# Loop the lagoons and output a continuous water quality profile for each
for(Lagoon in Lagoons$Lagoon){
  LagoonName = Lagoon
  ProfileContinuousWaterQuality(LagoonName)
}


```

# References

Jones T. 2017. Coastal Lagoon Vital Signs Monitoring Protocol for the Arctic Network: Standard Operating Procedures Version 1.0. National Park Service <https://irma.nps.gov/DataStore/Reference/Profile/2310242>

Jones T, Apsens S, Miller S, Robards M. 2018. Coastal lagoons vital signs monitoring protocol for the Arctic Network: Volume 1, report narrative, version 1.0. Natural Resource Report. NPS/ARCN/NRR—2018/1824. National Park Service. Fort Collins, Colorado[ https://irma.nps.gov/DataStore/Reference/Profile/2257596](https://irma.nps.gov/DataStore/Reference/Profile/2257596 "Click to go to Reference Profile page.")

Lawler J.P., Miller S.D., Sanzone D.M., Ver Hoef J., Young S.B. 2009. Arctic Network Vital Signs Monitoring Plan. Natural Resource Report. NPS/ARCN/NRR—2009/088. National Park Service. Fort Collins[ https://irma.nps.gov/DataStore/Reference/Profile/661340](https://irma.nps.gov/DataStore/Reference/Profile/661340 "Click to go to Reference Profile page.")
